Fractional Types: \\ First-Class Pattern Matching Clauses and Higher-Order Reversible Functions

### 2013-03-26 rc13/frac2.tex
abstract

\newcommand{\todo}[1]{\textbf{TODO:} #1}
\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
\newcommand{\amr}[1]{\textsc{Amr says:} #1}
\newtheorem{definition}{Definition}

\hyphenation{a-reas}

@@

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Fractional Types}
\authorinfo{Jacques Carette \and Amr Sabry}
           {McMaster University \and Indiana University}
           {carette@mcmaster.ca \and sabry@indiana.edu}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{Pi}} founded
  on type isomorphisms. Being restricted to first-order types limits the
  expressiveness of the language: it is not possible, for example, to
  abstract common program fragments into a higher-level combinator. In this
  paper, we introduce a higher-order extension of {{Pi}} based on the novel
  concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
  fractional type {{1/v}} represents \emph{negative} information. A
  function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
  representing the needed argument and {{v2}} representing the
  result. Fractional values are first-class: they can be freely propagated
  and transformed but must ultimately --- in a complete program --- be offset
  by the corresponding amount of positive information.
\end{abstract}

>>>>

\newtheorem{definition}{Definition}

@@

\titlebanner{}
\preprintfooter{}

\title{Fractional Types: \\
First-Class Pattern Matching Clauses and Higher-Order Reversible Functions}
\authorinfo{Jacques Carette \and Amr Sabry}
           {McMaster University \and Indiana University}
           {carette@mcmaster.ca \and sabry@indiana.edu}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{Pi}} founded
  on type isomorphisms. In subsequent work, we extended this language with
  \emph{fractional types}, the dual of conventional product types which, from
  an information-theoretic point of view, corresponds to the type of values
  with \emph{negative information}. The extended language was given an
  operational interpretation that hinted at two facts: fractional types can
  be used to define first-class pattern-matching clauses as well as
  higher-order reversible functions. Neither of these observations was backed
  by an appropriate denotational model, however. In this paper, we develop
  such a model by interpreting fractional types using functions mapping
  singleton types to the unit type. This interpretation requires that nearly
  all of the type constructs in {{Pi}} be re-interpreted as \emph{dependent
    types}. The framework, formalized in Agda, takes steps towards developing
  a computational interpretation of homotopy type theory using higher-order
  reversible languages.
\end{abstract}
